---
title: "4.SAPRC_Simulation"
author: "Felix Vo"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SAPRC exogenous variables in Biochemistry:

## Approximate PAN, HONO, and SAPRC Proxy
```{r, echo = FALSE}
# Load necessary packages
library(dplyr)
```

```{r}
data_path <- "C:/Users/felix/Desktop/CODING/felix's works/Air-Quality-Analysis-Project/data/final_cleaned_data.csv"
data <- read.csv(data_path)
```

## 1. Peroxyacetyl Nitrate (PAN) Approximation

### Formula (Using **Polynomial Regression** for C6H6):

$$
PAN_{proxy} = a \times NOx + b\times NOx^2 + c \times C6H6 + d\times C6H6^2 + e \times O_3 + f \times T
$$ 
Where:

\- **NOx (NO + NO₂)** → Major precursor for PAN.

-   **C6H6 (Benzene)** → A proxy for VOCs that contribute to PAN formation.

\- **C6H6² (Quadratic Term)** → Captures nonlinear VOC effects.

\- **O₃ (Ozone sensor value)** → PAN is strongly associated with photochemical smog.

\- **T (Temperature)** → Higher temperatures accelerate PAN formation.

```{r}

# Replace missing values and prepare data
data <- data %>%
  mutate(across(c(NOx.GT., C6H6.GT., PT08.S5.O3., T, RH), ~na_if(., -200))) %>%
  filter(!is.na(NOx.GT.), !is.na(C6H6.GT.), !is.na(PT08.S5.O3.), !is.na(T), !is.na(RH))

# Generate interaction and polynomial terms
data_model  <- data %>%
  mutate(
    # Polynomial terms
    NOx2 = NOx.GT.^2,
    C6H6_2 = C6H6.GT.^2,

    # Interaction terms
    NOx_RH = NOx.GT. * RH,
    NOx_T = NOx.GT. * T
  )

```
```{r}

# === 2. Estimate PAN Proxy Coefficients ===
model_pan <- lm(NO2.GT. ~ NOx.GT. + NOx2 + C6H6.GT. + C6H6_2 + PT08.S5.O3. + T, data = data_model)
cat("\n=== PAN Proxy Coefficients ===\n")
print(coef(summary(model_pan)))

```
```{r}
# === 2. PAN Proxy ===
# PAN_proxy = a * NOx + b * NOx^2 + c * C6H6 + d * C6H6^2 + e * O3 + f * T
data_model <- data_model %>%
  mutate(PAN_proxy = 0.225 * NOx.GT. - 0.000105 * NOx2 + 1.72 * C6H6.GT. - 0.0557 * C6H6_2 +
           0.0238 * PT08.S5.O3. - 0.202 * T + 37.04)
```



## 2. Nitrous Acid (HONO) Approximation

### Formula:

$$
HONO_{proxy} = f \times NOx + g \times RH + h \times (NOx\times Rh) + i \times T + j  \times(NO_x\times T)
$$ Where:

\- **NOx (NO + NO₂)** → HONO is formed from heterogeneous NO₂ reactions.

\- **RH (Relative Humidity)** → Surface chemistry enhances HONO formation under humid conditions.

\- **NOx Sensor (PT08.S3.NOx.)** → Used as an alternative for NO₂ levels.

\- **T (Temperature)** → Affects the equilibrium between HONO formation and photolysis.

```{r}
# === 1. Estimate HONO Proxy Coefficients ===
model_hono <- lm(NO2.GT. ~ NOx.GT. + RH + NOx_RH + T + NOx_T, data = data_model)
cat("=== HONO Proxy Coefficients ===\n")
print(coef(summary(model_hono)))

```

```{r}
# === 1. HONO Proxy ===
# HONO_proxy = f * NOx + g * RH + h * (NOx * RH) + i * T + j * (NOx * T)
data_model <- data_model %>%
  mutate(HONO_proxy = 0.328 * NOx.GT. - 0.499 * RH - 0.00204 * NOx_RH - 0.422 * T - 0.00246 * NOx_T + 98.30)
```

```{r}
# === 3. SAPRC Proxy ===
# SAPRC_proxy = a * NOx + b * C6H6 + c * C6H6^2 + d * O3 + e * T
data_model <- data_model %>%
  mutate(SAPRC_proxy = 0.121 * NOx.GT. + 2.885 * C6H6.GT. - 0.0891 * C6H6_2 +
           0.0300 * PT08.S5.O3. - 0.443 * T + 42.94)

# Optional: check correlation with NO2
cor(data_model$HONO_proxy, data_model$NO2.GT., use = "complete.obs")
cor(data_model$PAN_proxy, data_model$NO2.GT., use = "complete.obs")
cor(data_model$SAPRC_proxy, data_model$NO2.GT., use = "complete.obs")
```
```{r}
write.csv(data_model, "SAPRC_data.csv", row.names = FALSE)
```
```{r}
library(ggcorrplot)
data_numeric <- data_model %>% select(NO2.GT.,CO.GT., NOx.GT., PT08.S5.O3., T, RH, PAN_proxy, HONO_proxy, SAPRC_proxy )
cor_matrix <- cor(data_numeric, use = "complete.obs")

ggcorrplot(cor_matrix, method = "square", lab = TRUE, lab_size = 2.5)
```

