---
title: "Untitled"
author: "Felix Vo"
date: "2025-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulate SAPRC Outputs Using Our Dataset

------------------------------------------------------------------------

| **SAPRC Variable to Simulate** | **How We Can Estimate It?** |
|----|----|
| **PAN (Peroxyacetyl Nitrate)** | Estimate using NO₂, O₃, and Temperature. |
| **HONO (Nitrous Acid)** | Approximate from NO₂, RH, and Temperature. |
| **HNO₃ (Nitric Acid)** | Estimate using NO₂ and OH radical chemistry. |
| **OH Radicals** | Approximate from O₃, UV radiation (estimated from time of day). |

------------------------------------------------------------------------

atmospheric chemistry formulas to approximate SAPRC var.

### **Estimating PAN (Peroxyacetyl Nitrate) –\> Non linear**

**CH3C(O)OO + NO2 + M -\> PAN + M** PAN is formed from NO₂ and VOC oxidation and is affected by temperature:

$$
\text{PAN} = \alpha \cdot (\text{NO}_2 \cdot \text{O}_3) \cdot e^{\left( -\frac{E}{RT} \right)}
$$

-   $\alpha$ = Empirical constant = 0.01\
-   $E$ = Activation energy (from literature) = 500\
-   $R$ = Gas constant = 8.314\
-   $T$ = Temperature in Kelvin = C+ 273

### 2️⃣ **Estimating HONO (Nitrous Acid)**

HONO is a morning source of NO₂ and depends on NO₂ and humidity:

$$
\text{HONO} = \beta \cdot (\text{NO}_2) \cdot (\text{RH})
$$

-   $\beta$ = Empirical scaling factor = 0.05\
-   **RH** = Relative humidity

------------------------------------------------------------------------

## 📌 Step 3: Implement in R Using Your Data

Now, let’s **calculate PAN and HONO** using our existing dataset.

```{r}
# Load dataset
df <- read_csv("formated_data.csv")

# Convert Temperature to Kelvin
df$T_K <- df$T + 273.15

# Estimate PAN using an empirical formula
df$PAN <- 0.01 * (df$NO2.GT. * df$PT08.S5.O3.) * exp(-500 / (8.314 * df$T_K))
#### ____ OR ____
# Load necessary library
library(nlstools)

# Define nonlinear PAN model
pan_model <- nls(PAN ~ alpha * (NO2.GT. * PT08.S5.O3.) * exp(-E / (8.314 * (T + 273.15))),
                 data = df, start = list(alpha = 0.01, E = 500))

# Summary of the nonlinear model
summary(pan_model)


# Estimate HONO based on NO₂ and humidity
df$HONO <- 0.05 * (df$NO2.GT.) * (df$RH / 100)

# Display first rows
head(df[, c("PAN", "HONO")])
```

Step 4: Use Simulated SAPRC Outputs in SARIMAX

```{r}
# Convert NO₂ to time series
NO2_ts <- ts(df$NO2.GT., frequency = 24)

# Define exogenous variables (Simulated PAN and HONO)
exog_vars <- as.matrix(df[, c("PAN", "HONO")])

# Fit SARIMAX model
sarimax_model <- auto.arima(NO2_ts, xreg = exog_vars, seasonal = TRUE)

# Forecast using estimated PAN & HONO
future_exog <- tail(exog_vars, 24)
forecast_NO2 <- forecast(sarimax_model, xreg = future_exog, h = 24)

# Plot forecast
plot(forecast_NO2)

```

sources:: <https://en.wikipedia.org/wiki/Peroxyacetyl_nitrate?utm_source=chatgpt.com>

CREATE Create the New Exogenous Variable deriving it from the SAPRC model (Proxy SAPRC Variable)

Nitrogen dioxide (**NO₂**) is influenced by atmospheric chemistry, particularly the reactions:

-   **NO oxidation by O₃**:\
    $$ NO + O_3 \rightarrow NO_2 + O_2 $$\

-   **NO₂ photolysis (breakdown by sunlight (wavelength \< 400 nm))**:\
    $$ NO_2 + hv \rightarrow NO + O $$\

-   **NOₓ balance (total NO + NO₂)**:\
    $$ NO_x = NO + NO_2 $$

    We construct the proxy using:\
    ✅ **NOₓ (Total NO + NO₂):** Indicator of nitrogen oxides in the atmosphere.\
    ✅ **O₃ proxy (PT08.S5.O3.):** Since direct O₃ data is unavailable, we use sensor readings.\
    ✅ **CO:** A secondary pollutant that participates in photochemical reactions.\
    ✅ **Hour of the day:** Captures sunlight effects on NO₂ photolysis.

The estimated equation is: Fit a regression model to estimate the SAPRC proxy feature\
$$ SAPRC\_Proxy = \alpha \cdot NOx + \beta \cdot O3\_Sensor + \gamma \cdot CO + \delta \cdot Hour $$\
where **α, β, γ, δ** are coefficients estimated from data.

## 5️⃣ References

📚 **Seinfeld, J. H., & Pandis, S. N. (2016). Atmospheric Chemistry and Physics.**\
📚 **Carter, W. P. L. (2010). Development of the SAPRC-07 Chemical Mechanism.**\
📚 **Jacob, D. J. (1999). Introduction to Atmospheric Chemistry.** NO₂ photolysis (breakdown by sunlight) <https://acp.copernicus.org/articles/23/10413/2023/acp-23-10413-2023.pdf> <https://www.aeroqual.com/blog/meet-the-nitrogen-oxide-family> <https://en.wikipedia.org/wiki/NOx>

A simple linear approximation could be: SAPRC_proxy= a × NOx.GT.+ b × C6H6.GT. + c × PT08.S5.O3.+ d × T + e × RH Where the coefficients (a, b, c, d, e) can be determined using a regression approach by correlating these

.

To create an approximate proxy for SAPRC simulation outputs, we can derive a variable based on existing data. Since SAPRC models typically predict atmospheric chemistry transformations, we can approximate it using a combination of available pollutant and meteorological variables.

A possible proxy for SAPRC outputs could be a function of:

-   

-   **NOx (NOx.GT.)** and **VOC indicators (C6H6.GT., NMHC.GT.)** – since SAPRC models focus on photochemical reactions involving these pollutants.

-   

-   **Ozone-related sensor data (PT08.S5.O3.)** – as it is a key factor in NO₂ formation.

-   

-   **Temperature (T) and Humidity (RH)** – since atmospheric conditions significantly impact photochemical reactions.
